---
title       : Explore DNA
subtitle    : base frequency and k-mer enrichment
author      : Jianghan Qu
job         : 
framework   : io2012       # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## DNA: brief introduction

DNA, or deoxyribonucleic acid, is the hereditary material 
in humans and almost all other organisms. 
 

-  Human DNA consists of about 3 billion bases, and more than 99 percent of those bases are the same in all people.
-  The information in DNA is stored as a code made up of four chemical bases: adenine (A), guanine (G), cytosine (C), and thymine (T). 
-  The order, or sequence, of these bases determines the information available for building and maintaining an organism, similar to the way in which letters of the alphabet appear in a certain order to form words and sentences.

--- .class #id 

## Base frequency
- seqinr is an R package for biological sequences retrieval and analysis

```{r}
library('seqinr')
```

- Use an example of a short piece of DNA sequence, saved as a vector of characters

```{r}
dna <- s2c("acgggtacggtcccatcgaa")
```
- The base composition of this sequence is:

```{r}
count(dna, word = 1)
```

---

## K-mer frequency
```{r}
k <- 2; kmer_count  <- count(dna, wordsize = k)
```
```{r,echo=F}
library(ggplot2)
    kmer <- count(dna, wordsize=k)
      freq <- as.vector(kmer/(sum(kmer)-k+1))
      dat <- data.frame(Freq=freq, Nucleotide=names(kmer))
      ggplot(dat,aes(x=factor(Nucleotide), y=Freq)) + 
        geom_bar( stat="identity", aes(fill=Nucleotide)) +
        xlab(paste(k, "-mer",spe="")) + ylab("Proportion")
```

---&twocol  

## k-mer enrichment
*** =left
- Suppose the DNA sequence has length $L$
- Let $\# b$ be the number of occurrences of base $b\in \{A, T, G, C\}$ in the sequence 
- The expected occurrences of a k-mer $b_1\ldots b_k$ is : 
 $$L\times\prod_{i=1}^k p_{b_i}$$
 where 
 $p_{b_i} = \frac{\# b_i}{L}, b_i\in\{A, T, C, G\}.$
- Over-represented k-mers are above the diagnal line

*** =right
  ```{r, echo=FALSE}
    l <- length(dna)
    ct  <- count(dna, wordsize = 1)
    dat0 <- data.frame(Freq=as.vector(ct)/l, 
                       Nucleotide=names(ct),stringsAsFactors=F)
    row.names(dat0) <- dat0$Nucleotide

    dat$Exp <- 1.0
    for (i in 1:nrow(dat)){
      for (j in 1:k)
        dat$Exp[i] <- dat$Exp[i]*dat0[substr(dat$Nucleotide[i],j, j), 1]
    }
    library(ggplot2)
    ggplot(dat, aes(x=Freq, y=Exp, col=Nucleotide)) +
      geom_abline(intercept = 0, slope=1) +
      geom_point(shape=19, aes(fill=Nucleotide), cex=6) + 
      xlab("Observed proportion") +
      ylab("Expected proportion") +
      scale_x_continuous(limits = range(c(dat$Freq,dat$Exp)))+
      scale_y_continuous(limits = range(c(dat$Freq,dat$Exp)))
  ```

